Dual host-parasite transcriptomes of apicomplexan Eimeria falciformis and its natural mouse host

### 2016-10-07 manuscript/bmc_article.tex
manuscript updated

\textit{Eimeria} spp. infect the gut and are highly niche specific within parts and cell types of the intestines (cite?). There, these monoxenous parasites go through asexual (schizogony) and sexual reproduction which results in hosts releasing high numbers of resistant oocysts. (cite?) When a mouse ingests \textit{E. falciformis} oocysts, one sporulated oocyst releases eight infective sporozoites inside the host, which can infect epithelial crypt cells of the caecum and colon. Within the epithelium, so called merozoite stages form in several rounds of schizogony. (mesfin78) Parasite numbers increase drastically during schizogony. Oocysts are first detected in faeces on day six to seven after infection (cite?), and it peaks on day eight to nine p.i. Oocysts are not detectable after day 13-15 p.i. Oocysts form as merozoites differentiate into gametes, which fuse and form a zygote. Immature, unsporulated oocysts are shed into the environment where they mature into sporulated, infective oocysts.. (cite?)

@@

By applying dual RNA-seq and producing host and parasite transcriptomes from the same samples and tissue, we provide a first dataset which allows analysis of host and parasite mRNA profiles at several time-points post infection. Not only can we for the first time provide transcriptomes for the full parasite life cycle. In addition, having host transcriptome data from the same samples, we analyse biological (Gene Ontology, GO) processes in both species and discuss possible interactions of interest. This is particularly useful since the mouse is such a well annotated organism. 
Using the mouse model, we also had the possibility to include a mouse strain which lack a gene for maturation of T and B receptors, Recombination-activating gene 1, Rag1. Both strains were evaluated for re-infection after recovery. This design allows a full comparison of infections between immune competent mice and mice with severely limited adaptive immunity, and analysis of how these differences affect the parasite transcriptomes and reproductive success. We contextualize the parasite data by looking for enrichment of \textit{E. falciformis} specific genes, apicomplexa generic genes or widely conserved genes at different life cycle stages. We also correlate our RNA-seq data with published transcriptomes from \textit{E. tenella} and \textit{T. gondii}.

Taken together, the present study provides the first full life cycle transcriptome of an Eimeria spp. which naturally infects the research model M. musculus. In addition to providing a description of changes in the life cycle, we present the matching host transcriptomes and the host biological processes which take place during infection. This data is a source for in depth analyses and hypothesis testing of apicomplexan parasite infections in one of the best annotated model animals at hand. 
%\cite{koon,oreg,khar,zvai,xjon,schn,pond,smith,marg,hunn,advi,koha,mouse}

@@

We performed mRNA sequencing of caecum epithelial tissue from mice infected with
the apicomplexan parasite \textit{E. falciformis}. Oocysts and
sporozoites were included as "environmental" stages and processed
in vitro. To follow the life cycle of the parasite, we
compare different time-points after infection. We additionally used different mouse strains which display different immunocompetence in infection trials, measured by oocyst output in order to assess the influence of host immunocompetence on parasite development (Figure 1a and 1b). Immune competent NMRI mice were infected and sampled at three time-points post infection, p.i. We also infected mice lacking the Recombination activating gene 1, Rag1, (Rag1-/-) and compared them to the parental C57BL/6 mouse line. Rag1-/- mice lack mature B and T lymphocytes, which is taken as a proxy for absent adaptive immunity. All three mouse strains were used for infections of both naive mice and previously infected (and recovered) animals (onward referred to as "challenged" or "challenge infection").
Basic phenotyping (Figure 1a) showed differences in oocyst numbers between the immunocompetent (C57BL/6) and immunodeficient (Rag1-/-) host strains in naive and challenge infected mice. Immunocompetent NMRI mice were infected with a higher dosis of sporulated oocysts, and a drastic reduction of oocysts in faeces was seen in challenged mice compared to naive mice. Oocyst numbers in faeces peaked on days 8-9 andall mice had cleared the infection by day 14. Development and replication of \textit{E. falciformis} intestinal stages is reflected by the percentage of parasite reads sequenced per time-point post infection (Figure 1b) and this was confirmed by quantitative reverse transcription PCR (RT qPCR) of parasitic 18S (Figure 1c).
We thus use dual RNA-seq to analyse the life cycle of \textit{E. falciformis} under
the influence of host immune responses at early and late stages of infection. We used an experimental
design which allows to compare infections at 5 days post infection (dpi) for all experimental
conditions (NMRI, C57BL/6 and Rag1-/- mouse strains in naive and challenge infection). Additional time-points 3 dpi and 7 dpi were analysed from NMRI mice (Figure 1d).

\subsection*{Parasite and host dual transcriptomes can be assessed in parallel}
In this RNA-seq experiment, each replicate sample was enriched for caecum epithelial tissue and pooled 
from three mice. mRNA was extracted and sequencing libraries prepared. Two biological replicates were used for all but two conditions (with one and three replicates, respectively). Libraries were sequenced on several lanes of Illumina GAIIX (13 samples) and HiSeq machines (14 samples) and mapped to both mouse and parasite genomes simultaneously to avoid spurious assignments of reads in ultra conserved genomic regions. As samples and individual replicates were sequenced in batches to different depth and using different instrumentation (Table 1) we performed quality controls (additional files xyz). These confirm the absence of batch effects influencing analysis and quality of results.

Total numbers of sequenced reads as well as reads mapped to either the \textit{E. falciformis} genome or the mouse genome are indicated in Table 1 for all replicates. The number of total read mappings for individual replicates ranged from 25,362,739 (sample Rag\_1stInf\_0dpi\_rep1) to 139,749,046 (NMRI\_1stInf\_7dpi\_rep2). At the latest time-point samples, 7 dpi, the overall mRNA output of the sampled caecum tissue is dominated by parasite material with proportional mRNA abundance of 77\% (in NMRI\_1stInf\_7dpi\_rep2) and 92\% (in NMRI\_1stInf\_7dpi\_rep1). (Figure 1b).

\subsection*{Exclusion of samples with uncertain infection status}
A maximum of 92\% (sample NMRI\_1stInf\_7dpi\_rep1) and a minimum of 0.064\% (sample NMRI\_2ndInf\_3dpi\_rep2) of mapped reads could be assigned to the \textit{E. falciformis} genome in samples considered infected (Table 1). We excluded samples NMRI\_2nd\_3dpi\_rep1 and NMRI\_2nd\_5dpi\_rep2 due to low parasite contribution to the overall transcriptome. Technically, this exclusion made it possible to obtain read counts in agreement with a negative binomial distribution (see additional file x). It is also likely that the number of reads in the excluded samples would have been insufficient to fully normalise these datasets to those with the highest parasite contributions. From a biological point of view, both excluded samples are samples from challenge infection and we assume that the infection was cleared or reduced to a non-detectable level. One sample (NMRI\_1stInf\_0dpi\_rep1) was excluded because the uninfected control showed unexpected mapping of reads to the \textit{E. falciformis} genome. We consider the three excluded samples to display an uncertain state of infection.

\subsection*{The mouse transcriptome changes upon \emph{E. falciformis} infection}
Statistical testing for differential expression between infected and uninfected mice revealed changes in mRNA abundance becoming more pronounced (containing more genes) at later time-points post infection (Table 2). For mouse, 325 mRNAs were considered differently abundant (DA; FDR < 0.01) between controls and 3 dpi, 1,804 mRNAs between controls and 5 dpi and 2,711 mRNAs between controls and 7 dpi. This lead to a combined set of 3,453 unique genes responding to infection (Figure 2bi). DA mRNAs early in infection (3 dpi and 5 dpi) were not a strict subset of genes DA later in infection (7 dpi). Instead, the transcriptional profile of the mouse changes throughout the infection. Differences between controls and 7 dpi were in agreement with previously published microarry data. Fold-change data obtained from \textit{E. falciformis} infected mice at 6 dpi on Agilent microarrays (\cite{schmidt12}) or 7 dpi (our RNA-seq data) against uninfected controls show a strong correlation (Spearman's $\rho$ = 0.74; Figure 2a). Considering both biological differences in the experiments (e.g. exact time-points for samples), and technical differences between the two methods this comparison confirms the adequacy of using dual RNA-seq for assessing the host transcriptome.

\subsection*{Epithelial responses vary with mouse immune status}
A distinct response in early infection is indicated by the number of DA mRNAs between different time-points. To validate the pattern we performed hierarchical clustering on the (union of) mouse genes DA between different time-points post infection (Figure 2c). Three main sample clusters formed (dendrogram of columns at top of Figure 2c). Immune deficient Rag1-/- mice cluster with control samples. There is no clear distinction between infected and non-infected Rag1-/- samples, which confirms the immune deficiency phenotype in these mice. 
We identify a group of genes (cluster 4, Figure 2c) which changes transcriptional profile upon infection  in immune competent mice only. In immune compromised Rag1-/-, another group of genes (cluster 3, figure 2c) display the "infection profile" of immune competent mice also in control samples, indicating that these are mRNAs kept at a low abundance in healthy animals in a T and B cell dependent manner. Hence, these genes surprisingly appear to depend on functional T and B cells also in controls (cluster 3) and at the earliest time-point post infection (3 dpi, cluster 4). 
Gene Ontology, GO, terms enriched in cluster 3 are, e.g., "lipid metabolic process" and "protein intraciliary transport". Other enriched terms in cluster 3 are regulation of other metabolic processes and "blood coagulation" as well as terms containing "spinal cord", "axon" or "neuronal" regulation or development. (SI file x). We suggest that these are processes which are all intrinsically different in Rag1-/- mice compared to immune competent mice. The genes in cluster 4, which change only in immune competent mice, are highly abundant in controls including Rag1-/- and appear to be down-regulated upon infection by processes which depend on (mature) T and/or B cell activity, also as early as 3 dpi.

\subsection*{Renewal of infected epithelium appears heavily regulated}
Several terms enriched in cluster 4 are associated with wound healing and proliferation. 13 terms for cytokines as well as "negative regulation of viral (or inflammatory) response", "negative chemotaxis", "autophagy", "blood coagulation", "inositol phosphate-mediated signaling", and "positive regulation of calcineurin-NFAT" are enriched. mRNAS supporting these terms are less abundant in infected samples. Although speculative, several of these processes can be linked. Inositol signaling can lead to release of calcium and calcineurin-dependent translocation of NFAT to the nucleus and activation of its target genes in T cells, but also many other cell types (reviewed by macian05). Go enrichment also highlights regulation of transforming growth factor$\beta$, TGF$\beta$, epidermal growth factor, EGF, and tumor necrosis factor, TNF. TGF$\beta$ is important for wound healing in intestinal epithelium (beck03), and EGF regulates proliferation of epithelial cells and inhibits apoptosis (suzuki10). TNF is dose-dependent and can suppress inflammatory responses (noti10) and is reported to regulate proliferation of epithelial cells (kaiser97). Additionally, IL-1 and IL-6 are among the enriched GO terms in cluster 4. The IL-1 receptor (type I) is similar to Toll-like receptors and IL-1 induces innate immune responses in many cell types, and influences lymphocyte activity (dinarello09). IL-6 has been shown to support repair and inhibit apoptosis after epithelial wounding (kuhn14, probably through the Janus kinase, JAK, and signal transducer and activator of transcription STAT3 (pickert09). IL-6 is also known to be important for development of Th17 responses (ref in kuhn intro) which play an important role in responses to \textit{E. falciformis} (stange--). This analysis suggests that TGF$\beta$, TNF, EGF, IL-1 and IL-6 are main actors in the epithelial response to \textit{E. falciformis} infection and that the response is T and B cell dependent, also at early time-points. Enrichment tests suggest that hosts invest resources in intestinal healing also at early time-points of \textit{E. falciformis} infection and that functional T and B cells are needed for these responses. An alternative interpretation is that pathology is lower in Rag1-/- mice and that these responses are therefore not triggered in them. Based on unpunlished data from colleagues (talk to Stange), severe pathology in infected Rag1-/- mice makes this scenario unlikely.

\subsection*{Late infection is strongly enriched for adaptive immunity} 
The pronounced changes late in infection (7 dpi) reflect the expected onset of an adaptive immune response, undelined by enriched GO terms in gene clusters 1 and 5. Terms such as "antigen binding" and "immunoglobulin receptor binding" (molecular function, MF), and "immune system process", "adaptive immune response" and also "innate immune response" (biological process, BP) are highly enriched in cluster 5 and confirm an activated immune system and adaptive immune responses at this time-point. Among the same genes, natural killer cell regulation, JAK-STAT signalling, and IL-1 and interleukin-2, IL-2 production are enriched biological processes. IL-2 is one target of NFAT signaling and as mentioned above, JAK-STAT signaling can be induced by IL-6. It is likely that the enriched early responses in, e.g., NFAT and IL-6 regulation induce distinct mRNA abundace differences later in infection (7 dpi) and it is encouraging that these links are detected by the methods applied here. Taken together, clusters 3 and 4 highlight the difference between Rag1-/- mice and immune competent mice in this parasitic infection. Some genes are different also between control animals (Rag1-/- and NMRI, cluster 3), whereas in cluster 4 the profile is shared among all controls, including Rag1-/-. In both cases, the profile changes in immune competent mice but not in T and B cell deficient Rag1-/- mice. Using GO enrichment analysis, we show that regulation of IL-1, IL-6, TNF,TGF$\beta$ and EGF are regulated in both first and challenge infections and suggest wound healing as a probably function of this regulation, especially at 7 dpi.

@@

%% This might be relevant: The DMAP interaction domain of
%% UDP-GlcNAc:lysosomal enzyme
%% N-acetylglucosamine-1-phosphotransferase is a substrate recognition
%% module. Catalysis of the reaction: UDP-N-acetyl-D-glucosamine +
%% lysosomal-enzyme D-mannose = UMP + lysosomal-enzyme
%% N-acetyl-D-glucosaminyl-phospho-D-mannose

%% Go on here if you can...

%We detected further differences between infections of naive and
%challenge infected mouse hosts, especially at 7 dpi. These genes are
%characterized by...
%% We should't limit ourselfs to testing the clusters only! But
%% discuss the cluster first if you got difficulties coding the
%% enrichment tests.

\subsection*{Transcriptional differences in the parasite life cycle are independent of mouse immune status}
We performed statistical tests to evaluate significant differences (FDR$<$0.01) in mRNA abundance between different parasite life cycle stages, approximated by time post infection (Table 2). Between early time-points, 3 dpi and 5 dpi, 103 mRNAs were different, whereas between 3 dpi and 7 dpi 1399 mRNAs were DA, and between 5 dpi and 7 dpi 2084 mRNAs were DA (Figure 3a). This indicates that the major changes take place between 5 dpi and 7 dpi, and that variation is smaller between 3 dpi and 5 dpi. This motivated us to define 3 dpi and 5 dpi as "early infection" and 7 dpi as "late infection". Early and late infection samples were tested for DA compared to sporozoites and sporulated oocysts, resulting in 1697 and 3919 DA mRNAs, respectively.
%% Check venn vs table2 and change in ms + code
To evaluate this outcome further, we performed hierarchical clustering of (the union of) the DA mRNAs in the comparisons described above and applied GO enrichment analysis of gene clusters (annotations are inferred from orthologs in other \textit{Eimeria} spp. or \textit{T. gondii}). In the parasite transcriptome, we see no difference between infection in immune competent mice, or T and B cell deficient Rag1-/- mice, or between naive and challenge infected mice. This is surprising considering the measured differences in oocyst output in the same comparisons (Figure 1a), and the fact that these differences are visible in the mouse transcriptomes. On the parasite side, major patters instead seem to be determined by life cycle stages, independent of the host immune status. Distinct clusters of genes define early infection (3 dpi and 5 dpi) in which schizogony (asexual reproduction) takes place, and separately, late infection (7 dpi) in which it is assumed that gametocytes are present. Extracellular samples of sporozoites and oocysts also cluster separately and are defined by distinct gene clusters.

\subsection*{Early infection transcriptomes reflect parasite expansion}
GO enrichment of the major sporozoite defining cluster with high mRNA abundance in this stage (cluster 4) proposes that different biosynthesis processes are important, as well as "maintenance of protein location in cell". Possibly, the latter is due to control of microneme protein localization as sporozoites prepare for invasion. Biosynthesis terms could reflect preparation for asexual expansion. n reporting the genome of \textit{E. falciformis}, two fructose-bisphosphate aldolase, FBA, paralogs were inferred (heitlinger14). Based on previously reported data on \textit{T. gondii}, the authors speculate that localization of FBA is controlled in sporozoites in order to direct energy supply during the supposedly costly invasion. % check if FBA is in cluster 4
Our enrichment of control of protein localization support the idea that control over energy supply and probably other functions is important in the invasive stage.
In the following intracellular life cycle stages, measured at 3 dpi and 5 dpi, in which several rounds of schizogony (asexual replication) take place, mRNAs in cluster 6 are abundant in all early samples (except one, discussed below). Among these "early infection" genes in cluster 6, several GO terms (biological process) for biosynthetic acitivity are enriched, e.g., "ribosome biogenesis" and "cellular biosynthetic process", as well as terms for "gene expression" and RNA processing, including terms for tRNAs and ncRNAs. "Cellular amino acid catabolic process" is also enriched. In cluster 3, with a similar profile as cluster 6, no GO terms are enriched. Cluster 6 is distinct in early infection with a biosynthesis profile shared with invasive sporozoites. Replication and growth-related processes being enriched highlights the parasite's expansion in numbers on 3 dpi and 5 dpi, as supported both by previous knowledge about the life cycle and our increase in sequences from the parasite (Figure 1b).

>>>>

\textit{Eimeria} spp. infect the gut and are highly niche specific within parts and cell types of the intestines (cite?). There, these monoxenous parasites go through asexual (schizogony) and sexual reproduction which results in hosts releasing high numbers of resistant oocysts. (cite?) When a mouse ingests \textit{E. falciformis} oocysts, one sporulated oocyst releases eight infective sporozoites inside the host, which can infect epithelial crypt cells of the caecum and colon. Within the epithelium, so called merozoite stages form in several rounds of schizogony. (mesfin78) Parasite numbers increase drastically during schizogony, which is not completely synchronized and the exact number of schizogony rounds is not clear. Mesfin and Bellamy (1978) detected four schizont stages, but this has later been contradicted by our (? Richard?) unpublished data and others (???). Oocysts are first detected in faeces on day six to seven after infection (cite?), and it peaks on day eight to nine p.i. Oocysts are not detectable after day 13-15 p.i. Oocysts form as merozoites differentiate into gametes, which fuse and form a zygote. Immature, unsporulated oocysts are shed into the environment where they mature into sporulated, infective oocysts.. (cite?)

@@

By applying dual RNA-seq and producing host and parasite transcriptomes from the same samples and tissue, we provide a first dataset which allows analysis of host and parasite mRNA profiles at several time-points post infection. Not only can we for the first time provide transcriptomes for the full parasite life cycle. In addition, having host transcriptome data from the same samples, we analyse biological processes in both species and discuss possible interactions of interest. This is particularly useful since the mouse is such a well annotated organism. 
Using the mouse model, we also had the possibility to include a mouse strain which lack a gene for maturation of T and B cell receptors, Recombination-activating gene 1, Rag1. Both naive immune competent and T and B cell deficient mice were infected, and both groups were challenge infected after recovering from the first infection. This design allows a full comparison of infections between immune competent mice and mice with severely limited adaptive immunity, and analysis of how these differences affect the parasite transcriptomes and reproductive success. We contextualize the parasite data by testing for enrichment of \textit{E. falciformis} specific genes, apicomplexa specific genes or widely conserved genes in gene clusters which characterize different life cycle stages. We also correlate our RNA-seq data with published transcriptomes from the chicken parasite \textit{E. tenella} and \textit{T. gondii}.

Taken together, the present study provides the first full life cycle transcriptome of an Eimeria spp. which naturally infects the research model M. musculus. In addition to providing a description of changes in the life cycle, we present matching host transcriptomes and analyze the host biological processes which take place during infection. This data is a source for in-depth analyses and hypothesis testing of apicomplexan parasite infections in one of the best annotated model animals at hand. 
%\cite{koon,oreg,khar,zvai,xjon,schn,pond,smith,marg,hunn,advi,koha,mouse}

@@

We performed mRNA sequencing of caecum epithelial tissue from mice infected with
the apicomplexan parasite \textit{E. falciformis}. Oocysts and
sporozoites were included as "environmental" stages and processed
in vitro. To follow the life cycle of the parasite, we
compare different time-points after infection. We additionally used different mouse strains which display different immunocompetence in infection trials, measured by oocyst output, in order to assess the influence of host immunocompetence on parasite development (Figure 1a and 1b). Immune competent NMRI mice were infected and sampled at three time-points post infection, p.i. We also infected mice lacking Rag1 (Rag1-/-) and compared them to the parental C57BL/6 mouse line. Rag1-/- mice lack mature B and T lymphocytes, which is taken as a proxy for absent adaptive immunity. All three mouse strains were used for infections of both naive mice and previously infected (and recovered) animals (onward referred to as "challenged").
Basic phenotyping (Figure 1a) showed differences in oocyst numbers between the immunocompetent (C57BL/6) and immunodeficient (Rag1-/-) host strains in naive and challenge infected mice, as well as differences in amount of parasite 18S sequences determined by quantitative reverse transcription PCR (RT qPCR). These data suggest reduced parasite numbers and reduced reproductive success in immunocompetent, challenge infected mice. In contrast, no difference was seen between first and challenge infection in immunodeficient Rag1-/- mice. Immunocompetent NMRI mice were infected with a higher dosis of sporulated oocysts (150 in first infection and 1500 in challenge infection, which was required to see a response. Data not shown), and a drastic reduction of oocysts in faeces was seen in challenged mice. Oocyst numbers in faeces peaked on days 8-9 and all mice had cleared the infection by day 14. Asexual replication of \textit{E. falciformis} intestinal stages is reflected by the percentage of parasite reads sequenced per time-point post infection (Figure 1b) and this was confirmed by RT qPCR of parasitic 18S (Figure 1c).
We thus use dual RNA-seq to analyse the life cycle of \textit{E. falciformis} under the influence of different host immune capacity at early and late stages of infection. We used an experimental design which allows to compare infections at 5 days post infection (dpi) for all experimental conditions (NMRI, C57BL/6 and Rag1-/- mouse strains in naive and challenge infection). Additional time-points 3 dpi and 7 dpi were analysed from NMRI mice (Figure 1d).

\subsection*{Parasite and host dual transcriptomes can be assessed in parallel}
We purified host and parasite RNA from infected tissues, and we here demonstrate that even early in infection there is sufficient parasite material for detection by RNA-seq. Each replicate sample was enriched for caecum epithelial tissue and pooled from three mice. mRNA was extracted and sequencing libraries prepared. Two biological replicates were used for all but two conditions (with one and three replicates, respectively). Libraries were sequenced on several lanes of Illumina GAIIX (13 samples) and HiSeq machines (14 samples) and mapped to both mouse and parasite genomes simultaneously to avoid spurious assignments of reads in ultra conserved genomic regions. As samples and individual replicates were sequenced in batches to different depth and using different instrumentation (Table 1) we performed quality controls (additional files xyz). These confirm the absence of batch effects influencing analysis and quality of results. Total numbers of sequenced reads as well as reads mapped to either the \textit{E. falciformis} genome or the mouse genome are indicated in Table 1 for all replicates. The number of total read mappings for individual replicates ranged from 25,362,739 (sample Rag\_1stInf\_0dpi\_rep1) to 139,749,046 (NMRI\_1stInf\_7dpi\_rep2). At the latest time-points, 7 dpi, the overall mRNA output of the sampled caecum tissue is dominated by parasite material with proportional parasite mRNA abundance of 77\% (in NMRI\_1stInf\_7dpi\_rep2) and 92\% (in NMRI\_1stInf\_7dpi\_rep1). (Figure 1b) suggesting heavily infected tissues and large parasite numbers at this time-point.

\subsection*{Exclusion of samples with uncertain infection status}
Existing analysis methods of RNA-seq data (e.g. edgeR, DEseq) require certain distributions of data for reliable normalization and differential gene expression analysis. Low numbers of reads is a potential source of problematic, noisy distributions. We report a mapping to the parasite genome of maximum 92\% (sample NMRI\_1stInf\_7dpi\_rep1) and a minimum of 0.064\% (sample NMRI\_2ndInf\_3dpi\_rep2) in samples considered infected (Table 1). We excluded samples NMRI\_2nd\_3dpi\_rep1 (0.012\%) and NMRI\_2nd\_5dpi\_rep2 (0.023\%) due to low parasite contribution to the overall transcriptome. Technically, this exclusion made it possible to obtain read counts in agreement with a negative binomial distribution (see additional file x). It is also likely that the number of reads in the excluded samples would have been insufficient to fully normalise these datasets to those with the highest parasite contributions. From a biological point of view, both excluded samples are samples from challenge infection and it is likely that the infection had been cleared or reduced to a non-detectable level. One sample (NMRI\_1stInf\_0dpi\_rep1) was excluded because the uninfected control showed unexpected mapping of reads to the \textit{E. falciformis} genome. We consider the three excluded samples to display an uncertain state of infection.

\subsection*{The mouse transcriptome changes upon \emph{E. falciformis} infection}
We here show that upon infection with \textit{E. falciformis}, which induces weight loss and intestinal pathology in mice, the host transcriptome undergoes drastic changes with more than 3000 genes changing their mRNA profile.  Statistical testing for differential expression between infected and uninfected mice revealed changes in mRNA abundance becoming more pronounced (containing more genes) at later time-points post infection (Table 2). 325 mouse mRNAs were considered differently abundant (DA; FDR < 0.01) between controls and 3 dpi, 1,804 mRNAs between controls and 5 dpi and 2,711 mRNAs between controls and 7 dpi. This lead to a combined set of 3,453 unique genes responding to infection (Figure 2bi). DA mRNAs early in infection (3 dpi and 5 dpi) were not a strict subset of genes DA later in infection (7 dpi), which would be the case if the same genes are regulated throughout infection. Instead, the transcriptional profile of the mouse changes more fundamentally with other genes regulated late compared to early in infection. Our results are in agreement with previously published microarray data from the same species. Differences between our controls and 7 dpi samples were correlated with fold-change data obtained from \textit{E. falciformis} infected mice at 6 dpi on Agilent microarrays (\cite{schmidt12}). The data-sets show a strong correlation (Spearman's $\rho$ = 0.74; Figure 2a). Considering both biological differences in the experiments such as exact time-points for sampling, and technical differences between the two methods, this comparison confirms the adequacy of using dual RNA-seq for assessing the host transcriptome.

\subsection*{Epithelial responses vary with mouse immune status}
A distinct response in late infection is indicated by over 1600 DA mRNAs on 7 dpi (Figure 2b). To analyze this result further, we performed hierarchical clustering on the (union of) mouse genes DA between different time-points post infection (Figure 2b and 2c). Three main sample clusters formed (dendrogram of columns at top of Figure 2c). Immune deficient Rag1-/- mice, including infected samples, cluster with control samples. The failure to distinguish between infected and non-infected Rag1-/- samples, confirms the immune deficiency phenotype in these mice and suggest a strong influence of adaptive immune responses on sample and gene clustering. 
We identify a group of genes (cluster 4) which changes transcriptional profile upon infection in immune competent mice only, from high to low abundance. In immune compromised Rag1-/-, another group of genes (cluster 3) display the "infection profile" of immune competent mice also in control samples. These appear to be mRNAs which are kept at a low abundance in healthy animals in a T and B cell dependent manner. Hence, genes in cluster 3 surprisingly appear to depend on functional T and B cells also in uninfected mice (cluster 3) and at the earliest time-point post infection (3 dpi, cluster 4). 
Gene Ontology, GO, terms enriched in cluster 3 are, e.g., "lipid metabolic process" and "protein intraciliary transport". Other enriched terms are regulation of other metabolic processes and "blood coagulation" as well as terms containing "spinal cord", "axon" or "neuronal" regulation or development. (SI file x). We suggest that these are processes which are all intrinsically different in Rag1-/- mice compared to immune competent mice. The genes in cluster 4, which change only in immune competent mice, are highly abundant in controls including Rag1-/- and appear to be down-regulated upon infection by processes which depend on (mature) T and/or B cell activity, also as early as 3 dpi. Enriched GO biological processes include "signalling" and "cell communication" as well as terms containing "external stimuli" and "mating" and as described below, it appears that epithelial healing is induced .

\subsection*{Epithelium renewal is different depending on immune status}
Several terms enriched in cluster 4 are associated with wound healing and proliferation. 13 terms for cytokines as well as "negative regulation of viral (or inflammatory) response", "negative chemotaxis", "autophagy", "blood coagulation", "inositol phosphate-mediated signaling", and "positive regulation of calcineurin-NFAT" are enriched. mRNAS supporting these terms are less abundant in infected samples. Although speculative, several of these processes can be linked. Inositol signaling can lead to release of calcium and calcineurin-dependent translocation of NFAT to the nucleus and activation of its target genes in T cells, but also many other cell types (reviewed by macian05). GO enrichment also highlights regulation of transforming growth factor$\beta$, TGF$\beta$, epidermal growth factor, EGF, and tumor necrosis factor, TNF. TGF$\beta$ is important for wound healing in intestinal epithelium (beck03), and EGF regulates proliferation of epithelial cells and inhibits apoptosis (suzuki10). TNF is dose-dependent and can suppress inflammatory responses (noti10) and is reported to regulate proliferation of epithelial cells (kaiser97). Additionally, IL-1 and IL-6 regulation are among the enriched GO terms in cluster 4. The IL-1 receptor (type I) is similar to Toll-like receptors and IL-1 induces innate immune responses in many cell types, and influences lymphocyte activity (dinarello09). Both production and secretion appear to be downregulated by the genes is cluster 4. IL-6 has been shown to support repair and inhibit apoptosis after epithelial wounding (kuhn14, probably through the Janus kinase, JAK, and signal transducer and activator of transcription STAT3 (pickert09). IL-6 is also known to be important for development of Th17 responses (ref in kuhn intro) which play an important role in responses to \textit{E. falciformis} (stange--). This analysis suggests that TGF$\beta$, TNF, EGF, IL-1 and IL-6 are main actors in the epithelial response to \textit{E. falciformis} infection and that the response is T and B cell dependent, also at early time-points. Enrichment tests suggest that hosts invest resources in intestinal healing and possible regulatory functions (IL-1 down) also at early time-points of \textit{E. falciformis} infection. The difference here between immunocompetent and immunodeficient Rag1-/- mice suggest that functional T and B cells are needed for these responses. An alternative interpretation is that pathology is lower in Rag1-/- mice and that these responses are therefore not triggered in them. However, based on unpublished data from colleagues (talk to Stange?), severe pathology seen in infected Rag1-/- mice makes this scenario unlikely.

\subsection*{Late infection is strongly enriched for adaptive immunity} 
The pronounced changes late in infection (7 dpi) reflect the expected onset of an adaptive immune response, undelined by enriched GO terms in gene clusters 1 (also abundant in early samples) and 5. Terms such as "antigen binding" and "immunoglobulin receptor binding" (molecular function, MF), and "immune system process", "adaptive immune response" and also "innate immune response" (biological process, BP) are highly enriched in cluster 5 and confirm an activated immune system and adaptive immune responses at this time-point. Among the same genes, natural killer cell regulation, JAK-STAT signalling, and IL-1 and interleukin-2, IL-2 production are enriched biological processes. IL-2 is one target of NFAT signalling and as mentioned above. JAK-STAT signalling can be induced by IL-6. It is likely that the enriched early responses in, e.g., NFAT and IL-6 regulation induce distinct mRNA abundace differences later in infection (7 dpi) and it is encouraging that these links are detected by the methods applied. Taken together, the clustering patterns of immunocompetent versus immunocompromised hosts suggest a strong influence of adaptive immunity on hierarchical clustering, and therefore on the major biological processes which characterize the infection. Using GO enrichment analysis, we additionally show that regulation of IL-1, IL-6, TNF, TGF$\beta$ and EGF are regulated in both first and challenge infections and suggest wound healing as a probable function of this regulation, especially at 7 dpi.

@@

\subsection*{Transcriptional differences in the parasite life cycle are independent of mouse immune status}
We present data which supports that \textit{E. falciformis} is "transcriptionally blind" to the host immune status, since comparisons between differently immune-competent mice return no significantly different mRNAs. We performed statistical tests to evaluate significant differences (FDR$<$0.01) in mRNA abundance between different parasite life cycle stages, approximated by time post infection (Table 2). Between early time-points, 3 dpi and 5 dpi, 103 mRNAs were different, whereas between 3 dpi and 7 dpi 1399 mRNAs were DA, and between 5 dpi and 7 dpi 2084 mRNAs were DA (Figure 3a). This indicates that the major changes take place between 5 dpi and 7 dpi, and that variation is smaller between 3 dpi and 5 dpi. This motivated us to define 3 dpi and 5 dpi as "early infection" and 7 dpi as "late infection". Early and late infection samples were tested for DA compared to sporozoites and sporulated oocysts, resulting in 1697 and 3919 DA mRNAs, respectively.
%% Check venn vs table2 and change in ms + code
To evaluate this outcome further, we performed hierarchical clustering of (the union of) the DA mRNAs in the comparisons described above and applied GO enrichment analysis of gene clusters (annotations are inferred from orthologs in other \textit{Eimeria} spp. or \textit{T. gondii}). In the parasite transcriptome, we see no difference between infection in immune competent mice, or T and B cell deficient Rag1-/- mice, or between naive and challenge infected mice. This is surprising considering the measured differences in oocyst output in the same comparisons (Figure 1a), and the fact that these differences are visible in the mouse transcriptomes. Major patters in the parasite transcriptome instead seem to be determined by life cycle stages, independent of the host immune status. Distinct clusters of genes define early infection (3 dpi and 5 dpi) in which schizogony (asexual reproduction) takes place, and separately, late infection (7 dpi) in which it is assumed that gametocytes are present. Extracellular samples of sporozoites and oocysts also cluster separately and are defined by distinct gene clusters.

\subsection*{Early infection transcriptomes reflect parasite expansion}
Our data suggest ATP production as a dominant feature in sporozoites, followed by biosynthetic processes in the schizogony stages that follow upon invasion. GO enrichment of the major sporozoite defining cluster with high mRNA abundance in this stage (cluster 4) returns different biosynthesis processes. Metabolic pathway analysis (ToxoDB) of the same genes reveals several pathways which suggest that ATP production takes place, including fatty acid degradation, oxidative phosphorylation and  valine, leucine and isoleucine degradation (SI xx). The latter degradation of branched chain amino acids pathway is enriched by genes which have orthologs in \textit{T. gondii}, where they are used to utilize amino acids for energy supply (oppenheim14). In addition, the invasive stage is characterized by "maintenance of protein location in cell" and similar terms. Possibly, this is due to control of microneme protein localization as sporozoites prepare for invasion. In the following life cycle stages, measured at 3 dpi and 5 dpi, in which several rounds of schizogony take place, mRNAs in cluster 6 are abundant in all early samples (except one, discussed below). Among these "early infection" genes in cluster 6, several GO terms (biological process) for biosynthetic acitivity are enriched, e.g., "ribosome biogenesis" and "cellular biosynthetic process", as well as terms for "gene expression" and RNA processing, including terms for tRNAs and ncRNAs. "Cellular amino acid catabolic process" is also enriched. In the small gene cluster 3, no GO terms are enriched. Cluster 6 is distinct in early infection with a biosynthesis profile similar to that of invasive sporozoites. Replication and growth-related processes being enriched highlights the parasite's expansion in numbers on 3 dpi and 5 dpi, as supported both by previous knowledge about the life cycle and our increase in parasite derived sequences (Figure 1b). The mRNAs supporting the processes described here all have a low abundance late in infection, on 7 dpi. This can be explained by a switch from the early asexual expansion towards sexual reproduction on the time-point one day before oocyst outputs peak.
